Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-01 ms/ms.Rnw
finished first draft of phylo figure
adding growth form comparison

I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). \Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny. For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position of a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining \Sexpr{nNotInPhy2} missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

I excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species. The R code accompanying this paper documents these decisions with citations to the relevant literature.

@@

>>>>

I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Niinemets_Valladares_2006, Bartelheimer_Poschlod_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} as measures of light habitat. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Ellenberg light indicator values, hereafter abbreviated \el, have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}).

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species were classified as shrubs (plant height less than \Sexpr{round(shrub_cutoff, 0)} m) or trees (plant height greater than \Sexpr{round(shrub_cutoff, 0)} m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}. [NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]

I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). \Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny. For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position of a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining \Sexpr{nNotInPhy2} missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

I excluded data on hyrdrophytes (\Sexpr{nHydrophyte} species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. I also excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo}). The R code accompanying this paper documents these decisions with citations to the relevant literature.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{CAPTION} 
\label{fig:phylo}
\end{figure}

### 2017-10-01 ms/ms.tex
adding growth form comparison

I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). 14 species (3.5\%) in the dataset were not present in the phylogeny. For 8 of these species, I used the position of a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo.pdf}}
\caption{CAPTION} 
\label{fig:phylo}
\end{figure}

>>>>

I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Niinemets_Valladares_2006, Bartelheimer_Poschlod_2016, Shipley_etal_2017}, I used Ellenberg light indicator values \citep{Ellenberg_1974} as measures of light habitat. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Ellenberg light indicator values, hereafter abbreviated \el, have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}).

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \citeauthor{Muir_2015} (\citeyear{Muir_2015}), `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species were classified as shrubs (plant height less than 4 m) or trees (plant height greater than 4 m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Salguero-Gomez_etal_2016}. [NOTE: SUPPLEMENTAL FIGURE OF OVERLAP BETWEEN THESE]

I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). 14 species (3.5\%) in the dataset were not present in the phylogeny. For 8 of these species, I used the position of a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{CAPTION} 
\label{fig:phylo}
\end{figure}
